Future Cosmic Microwave Background delensing with galaxy surveys.

### 2017-07-03 Notes/galaxy_delensing.tex
text: section 4

\subsection{CMB-S3 Era}
CMB polarization measurements are rapidly improving. Indeed the next generation of ground-based telescopes has been already deployed, and data are currently being taken.
We model the CMB S3 stage with a deep experiment with a level of noise of \textcolor{red}{$3\mu$K-arcmin} with a \textcolor{red}{1 arcmin} beam. 
We assume here that a deep CMB experiment and an overlapping high resolution one will be able to combine their measured modes.
For this reason, we will also assume this level of noise for the internal noise reconstruction. With this experimental setup we can push the angular scale range used to calculate the Fisher matrix to $50<\ell<3000$ even if most of the high $\ell$ scale do not contribute to the constraints.

Not only will CMB experiments improve shortly: eventually DESI will start taking data. For this reason, we add DESI to the previously mentioned LSS tracers. 
The correlation attainable using generation 3 experiments is shown in \reffig{corrS3}.
An interesting finding is that DESI will be less efficient (removing 10\% of the power) than a DES-like survey despite the fact that it can probe slightly higher redshift. The reason is that, for the broad CMB kernel, spectroscopic redshift accuracy is not needed and the lower shot noise in DES play a significant role in the delensing efficiency. Unfortunately, adding DESI will only bring the power removed using LSS fro 36 to 41\%.
Indeed, even in the near future, CIB will still play the dominant role among LSS.
Contrary to current experiments, th CMB internal reconstruction will dominate the correlation with the lensing potential up to $\ell\simeq 550$.

%This generation of experiments will start exploring the small scales B-modes thus constraining, even if loosely, the value of $n_{t}$. 
The improvement in cosmological parameters is summarized in \reftab{S3}. 
The main difference from the current generation of experiments is that in the near future CMB alone will be able to improve constraint on $r$ by a factor of 2 through internal delensing in the ideal case of no instrumental noise.

Despite this compared to CMB alone, galaxy surveys will still be able to remove an additional $20\%$ of power and to improve our constraint on the null hypothesis ($r=0$) by $15\%$. Note that apart from the addition of DESI, which plays a secondary role, the tracers used here are already available for the current generation. The ongoing effort can then allow this improvement even for the next generation of experiments. 
The combination of CMB and galaxy surveys will be able to improve the constrain on the null hypothesis of no primordial waves by a factor 1.8 or to improve the constraint of a possible detection (r=0.12) by 30\%.

\relook{something about $n_t$}

@@

@@

An ambitious program for a generation 4 ground CMB experiment is currently under planning \cite{abazajian:2016}. Moreover, satellite experiments have been proposed.

The correlation attainable using current experiments is shown in \reffig{corrs4},
In this futuristic but reasonable scenario, we add to the available large scale structures tracers both 21 cm surveys like SKA and large scale structures like LSST and Euclid.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=1.]{../images/S4_scenario_no_ska.pdf}
\caption{Correlation factor. Same as \reffig{corrnow} but for stage 4 experiments without SKA. }
\label{fig:corrs4}

\end{center}
\end{figure}

@@

\begin{table}
\caption{$\alpha$: Gen 4 experiments}
  \begin{tabular}{ | c | c  |  c |}
\hline
Surveys & $\alpha(r)$ & $\alpha(n_{t})$\\ \hline \hline
Euclid & 1.53 & \\ \hline
LSST & 1.65  &\\ \hline
SKA & 4.35  &\\ \hline
CMB S4 & 3.7  &\\ \hline
LSS+CMB & 6.3  &\\ \hline

  \end{tabular}
  \label{tab:S4}
\end{table}

>>>>

\subsection{CMB-S3 Era}
CMB polarization measurements are rapidly improving. Indeed the next generation of ground-based telescopes has been already deployed, and data are currently being taken.
We model the CMB S3 stage with a deep experiment with a level of noise of \textcolor{red}{$3\mu$K-arcmin} with a \textcolor{red}{1 arcmin} beam. 
We assume here that a deep CMB experiment and an overlapping high resolution one will be able to combine their measured modes.
For this reason, we will also assume this level of noise for the internal noise reconstruction. With this experimental setup we can push the angular scale range used to calculate the Fisher matrix to $50<\ell<3000$ even if most of the high $\ell$ scale do not contribute to the constraints.
Not only will CMB experiments improve shortly: eventually DESI will start taking data. For this reason, we add DESI to the previously mentioned LSS tracers.

The correlation factor attainable using generation 3 experiments is shown in \reffig{corrS3}.
An interesting finding is that DESI will be less efficient (removing 10\% of the power) than a DES-like survey (14\%) despite the fact that it can probe slightly higher redshift. The reason is that, for the broad CMB kernel, spectroscopic redshift accuracy is not needed and the lower shot noise in DES play a significant role in the delensing efficiency. Unfortunately, adding DESI will only bring the power removed using LSS fro 36 to 41\%.
Indeed, even in the near future, CIB will still play the dominant role among LSS.
Contrary to current experiments, th CMB internal reconstruction will dominate the correlation with the lensing potential up to $\ell\simeq 550$.

%This generation of experiments will start exploring the small scales B-modes thus constraining, even if loosely, the value of $n_{t}$.

@@

The improvement in the constraint on $r$ is summarized in \reftab{S3}. 
Given the improvement in the noise of the high resolution experiment, CMB alone will be able to improve constraint on $r$ by a factor of 2 through internal delensing in the ideal case of no instrumental noise.
Despite this, adding galaxy survey will still lead to a decent improvement with a fairly low effort.
Indeed galaxy surveys will still be able to remove an additional $20\%$ of power from $C_{\ell}^{BB,\mathrm{lens}}$ and to improve our constraint on the null hypothesis ($r=0$) by $15\%$. Note that apart from the addition of DESI, the tracers used here are already available for the current generation. The ongoing effort can then lead to a significant improvement even for the next generation of experiments. 
Finally, the combination of CMB and galaxy surveys will be able to improve the constrain on the null hypothesis of no primordial waves by a factor 1.8 or to improve the constraint of a possible detection (r=0.12) by 30\%.
As expected the improvement on $n_{t}$ is similar to the one in $r$ since they are both proportional to the variance of the measured B-modes. We tested this in the scenario $r=0.12$ where we additionally impose the consistency relation $n_{t}^{\rm{fid}}=-r^{\rm{fid}/8}$. 
It is important to note that, independently from the level of delensing, for CMB 3G experiments this statistical error will still be several times bigger than the fiducial value.

@@

An ambitious program for a generation 4 ground CMB experiment is currently under planning \cite{abazajian:2016}. 
Moreover, satellite and balloon experiments have been proposed and have the potential to extend the accessible B-mode measurement to the largest scales.
Here we follow \cite{abazajian:2016} and assume a CMB-S4 ground experiment with  with a level of noise of \textcolor{red}{$1\mu$K-arcmin} with a \textcolor{red}{1 arcmin} beam.
By the mid 2020s several next generation of LSS survey will be online. Here, as an example for future optical galaxy surveys we add LSST \cite{lsst-science-collaboration:2009} to the CMB lensing tracers. We test that other experiments like Euclid \cite{laureijs:2011} and WFIRST \cite{spergel:2013} will have similar delensing performance. Even if they will observe similar part of the sky, because of the different  strength and weakness of this experiments  
it will still important to combined them efficiently.
We also consider a radio continuum survey modeled following the SKA planned experiment. Radio continuum observation of the 21 cm line is in its early stage and several experimental and analysis challanges 
need to be overcome. However this technique has the potential to map LSS with relatively low noise up to redshift $z=6$.

The correlation factor with CMB lensing of stage 4 experiments is shown in \reffig{corrs4}.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=1.]{../images/S4_scenario_no_ska.pdf}
\caption{Correlation factor. Same as \reffig{corrnow} but for stage 4 experiments without SKA. }
\label{fig:corrs4}
\end{center}
\end{figure}

A tomographically binned LSST-like experiment will be a very efficient CMB lensing tracer. Indeed it will be more than 70\% correlated with CMB lensing for all the scales $\ell<600$ a performance similar to a Stage 3 CMB internal reconstruction.
SKA very new but potentially amazing
With SKA and LSST, for the first we will have LSS tracers way more efficient than the CIB.
Even if galaxy survey will improve dramatically the CMB internal reconstruction will still be the main source of delensing. This, having a perfect kernel overlap with the true lensing potential we will benefit for the very low level of noise of CMB S4 experiments.
CMB S4 delensing will probably be limited by small secondary effect like foregrounds contamination, missing modes due to filtering etc.

\reftab{S4} shows the improvement on the inflationary constraint.

@@

\begin{table}
\caption{$\alpha$: Gen 4 experiments}
  \begin{tabular}{ | c | c  |  c |}
\hline
Surveys & $\alpha(r)$ & $\alpha(n_{t})$\\ \hline \hline
Euclid & 1.53 & \\ \hline
LSST & 1.65  &\\ \hline
SKA & 4.35  &\\ \hline
CMB S4 & 3.7  &\\ \hline
LSS+CMB & 6.3  &\\ \hline
\end{tabular}
\label{tab:S4}
\end{table}

\begin{table}
\caption{$\alpha$: Gen 4 experiments}
  \begin{tabular}{ | c | c  |  c |}
\hline
Surveys & $\alpha(r)$ & $\alpha(n_{t})$\\ \hline \hline
Euclid & 1.53 & \\ \hline
LSST & 1.65  &\\ \hline
SKA & 4.35  &\\ \hline
CMB S4 & 3.7  &\\ \hline
LSS+CMB & 6.3  &\\ \hline
\end{tabular}
\label{tab:S4}
\end{table}
